### 2014-12-01 text/Abstract.tex
Proofreading chapters 1 and 2, reorder chapter 6

The key point made by this thesis is the realization that execution environment or
\emph{context} is fundamental for writing modern applications and that programming 
languages should provide abstractions for programming with context and verifying how 
it is accessed.

@@

Our first contribution is the discovery of the connection between the above examples and
their novel presentation in the form of calculi (\emph{coeffect systems}). The presented type 
systems and formal semantics highlight the relationship between different notions of context. 
Our second and third contributions are two unified coeffect calculi that capture commonalities 
in the presented examples. In particular, our \emph{flat coeffect calculus} models languages 
with contextual properties of the execution environment and our \emph{structural coeffect 
calculus} models languages where the contextual properties are attached to the variable usage.

>>>>

The key point made by this thesis is the realization that an execution environment or
a \emph{context} is fundamental for writing modern applications and that programming 
languages should provide abstractions for programming with context and verifying how 
it is accessed.

@@

Our first contribution is the discovery of the connection between the above examples and
their novel presentation in the form of calculi (\emph{coeffect systems}). The presented type 
systems and formal semantics highlight the relationship between different notions of context. 
Our second and third contributions are two unified coeffect calculi that capture commonalities 
in the presented examples. In particular, our \emph{flat coeffect calculus} models languages 
with contextual properties of the execution environment and our \emph{structural coeffect 
calculus} models languages where the contextual properties are attached to the variable usage.
We close the thesis with a \emph{unified coeffect calculus} that captures the two notions
using a single parameterized system.

### 2014-12-01 text/Chapter01.tex
Proofreading chapters 1 and 2, reorder chapter 6

\item \textbf{Resources \& data availability.} When creating a mobile application, the program may
  (or may not) be granted access to device capabilities such as GPS sensor, social updates or battery
  status. We would like to know which of the capabilities are required and which are optional
  (\ie~enhance the user experience, but there is a fallback strategy). Equally, on the server-side,
  we might have access to different database tables and other information sources.
\end{itemize}

\noindent
Most developers do not perceive the above as programming language flaws -- they are simply common
programming problems (at most somewhat annoying and tedious) that have to be solved. However, this
is because it is not apparent that a suitable language extension could make the above problems
significantly easier to solve. As the number of distinct contexts and their diversity increases,
these problems will become even more commonplace.

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
\kvd{for}~\ident{header},\ident{value}~\kvd{in}~\ident{header}~\kvd{do}\\[-0.25em]
\hspace{2em}\kvd{match}~\ident{header}~\kvd{with}\\[-0.25em]
\hspace{2em}|~\str{accept} \rightarrow~ \ident{req.Accept} \leftarrow \ident{value}\\[-0.25em]
%
\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_USERAGENT}\\[-0.25em]
\hspace{2em}|~\str{user-agent} \rightarrow~ \ident{req.UserAgent} \leftarrow \ident{value}\\[-0.25em]
\prepk{\#else}\\[-0.25em]
\hspace{2em}|~\str{user-agent} \rightarrow~
  \ident{req.Headers}.[\,\ident{HttpHeader.UserAgent}\,] \leftarrow \ident{value}\\[-0.25em]
\prepk{\#endif}\\[-0.25em]
%
\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_REFERER}\\[-0.25em]
\hspace{2em}|~\str{referer} \rightarrow~ \ident{req.Referer} \leftarrow \ident{value}\\[-0.25em]
\prepk{\#else}\\[-0.25em]
\hspace{2em}|~\str{user-agent} \rightarrow~
  \ident{req.Headers}.[\,\ident{HttpHeader.Referer}\,] \leftarrow \ident{value}\\[-0.25em]
\prepk{\#endif}\\[-0.25em]
%
\hspace{2em}|~\ident{other} \rightarrow~ \ident{req.Headers}.[\,\ident{other}\,] \leftarrow \ident{value}\\[-0.25em]
\end{array}
\end{equation*}

\vspace{1em}
\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}}
\vspace{-1.5em}
\caption{Conditional compilation in the HTTP module of the F\# Data library}
\label{fig:introduction-context-http}
\end{figure}

@@

As an alternative to conditional compilation, developers can use dynamic loading. For example,
on Android, programs can access API from higher level platform dynamically using techniques like
reflection and writing wrappers. This is even more error prone. As noted in an article\footnote{Retrieved 
from: \url{http://android-developers.blogspot.com/2009/04/backward-}\\\url{compatibility-for-android.html}}
introducing the technique ``Remember the mantra: if you haven't tried it, it doesn't work.'' 
Again, it would be reasonable to expect that statically-typed languages could provide a better solution.

@@

Another example related to the previous one is when libraries use meta-programming techniques
(such as LINQ \cite{app-linq} or F\# quotations \cite{app-fsharp-metaprog}) to translate code
written in a subset of a host language to some other target language, such as SQL, OpenCL or
JavaScript. For database access, this is a recently developed technique replacing embedded SQL
discussed in the introduction, but it is a more generally important technique for programming
in heterogeneous environments. It lets developers targets multiple runtimes that have limited 
execution capabilities.

@@

When cross-compiling to a limited environment, we cannot always support the full source language.
The example with LINQ and SQL demonstrates the importance of this problem. As of March 2014,
Google search returns 11800 results for the message above and even more results (44100) for a
LINQ error message \emph{``Method X has no supported translation to SQL''} caused by a similar
limitation.

@@

The previous two examples were related to the non-existence of some library functions in a different
execution environment. Another common factor was that they were related to the execution context of 
the whole program or a function scope. However, contextual properties can also be associated with
a specific variables.

For example, consider the following code sample that accesses a database by building a SQL query
using string concatenation. For the purpose of the demonstration, this example does not use LINQ, 
but an older approach with a parameterized SQL query written as a string:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{query} = \ident{sprintf}~\str{SELECT * FROM Products WHERE Name='\%s'}~\ident{name}\\[-0.25em]
\kvd{let}~\ident{cmd} = \kvd{new}~\ident{SqlCommand}(\ident{query})\\[-0.25em]
\kvd{let}~\ident{reader} = \ident{cmd.ExecuteReader}()
\end{array}
\end{equation*}
%
The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
\cite{app-sql-injection} (an attacker could enter \str{'; DROP TABLE Products -{}-} as their 
\ident{name} and delete the database table ``Products''). For this reason, most libraries discourage 
building SQL commands by string concatenation, but there are still many systems that do so.

The example demonstrates a more general property. Sometimes, it is desirable to track additional
meta-data about variables that are in some ways special. Such meta-data can determine how the variables
can be used. Here, \ident{name} comes from the user input. This information about the value should
be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
not directly contain user input (in an unchecked form).

@@

In the security context, such marking of values (but at run-time) is called \emph{tainting} 
\cite{app-tainting-sql}, but the technique is a special case of more general \emph{provenance}
tracking. This can be useful when working with data in other contexts. For example, 
data jounralsts might want to propagate meta-data about the quality and the information source -- 
is the source trustworthy? Is the data up-to-date? Such meta-data could propagate to 
the result and tell us important information about the calculated results.

@@

The information about access patterns can be used to efficiently compile the computation
by preallocating the necessary space (as we know which sub-range of the array might be accessed).
It also allows better handling of boundaries. For example, to simplify wrap-around behaviour we 
could pad the \ident{input} with a known number of elements from the other side of the array.

@@

The four examples presented in the previous section cover different kinds of \emph{context}.
The context includes notions such as execution environment, capabilities provided by the 
environment or input and meta-data about the input.

@@

\vspace{1em}
\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}}
\vspace{-1.5em}
\caption{News reader implemented in a context-aware language}
\label{fig:introduction-sample}
\end{figure}

@@

The approach advocated in this thesis is to track information about context requirements using
the type system. To make this practical, the system needs to provide at least partial support for
automatic type inference, as the information about context requirements makes the types more complex.
An inspiring example might be the F\# support for units of measure \cite{app-units} -- the user has
to explicitly annotate constants, but the rest of the information is inferred automatically.

Furthermore, integrating contextual information into the type system can provide information for
modern developer tools. For example, many editors for F\# display inferred types when placing mouse
pointer over an identifier. For \ident{fetchLocalNews}, the tip could appear as follows:
%
\definecolor{tooltipy}{rgb}{0.98,0.97,0.76}
\definecolor{tooltipb}{rgb}{0.7,0.7,0.7}
\newcommand{\introfunc}[3]{ \ident{#1 \hspace{-1em}\footnotesize{@} \{ #2 \}} $\rightarrow$ \ident{#3}  }
\newcommand{\introvar}[2]{ \ident{#1 \hspace{-1em}\footnotesize{@} #2 } }
\vspace{0.5em}
\begin{equation*}
\hspace{5em}
\begin{array}{l}
\ident{\textbf{fetchLocalNews}}\\
\hspace{1em} \fcolorbox{tooltipb}{tooltipy}{
	\introfunc{unit}{gps, rpc}{(news~list)~async}
}
\end{array}
\end{equation*}
\vspace{0.5em}

@@

\item The \ident{fetchNews} function requires database access and so it can only run on the server-side
  (or on a thick client with local copy of the database, such as a desktop computer with an offline mode).

@@

In this scenario, the main benefit of using an integrated context-aware language would be the
ability to design appropriate abstractions using standard mechanisms of the language. For
cross-compilation, we can structure code using functions, rather than relying on \prepk{\#if}
directives. Similarly, the splitting between client-side, server-side and shared code can be
done using ordinary functions and modules -- rather than having to split the application into
separate independent libraries or projects.

@@

\paragraph{Type systems.}
A type system is a form of static analysis that is usually specified by \emph{typing judgements}
such as $\Gamma \vdash e : \tau$. The judgement specifies that, given some variables described by
the context $\Gamma$, the expression $e$ has a type $\tau$.
The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an
expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int},
\ident{float}, or even \ident{string}, depending on the types of the variables. For example, 
the following is a valid typing judgement in F\#:
%
\begin{equation*}
x\!:\!\ident{int},~y\!:\!\ident{int} \vdash x+y : \ident{int}
\end{equation*}
%
This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
be \ident{int}. In F\#, the expression would also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
but not, for example, in a context where $x$ has a type \ident{int} and $y$ has a type \ident{string}.

@@

The systems for tracking of context requirements developed in this thesis are inspired by the idea
of effect systems. To demonstrate our approach, consider the following call from the sample program
shown earlier -- first using standard ML-like type system:
%
\begin{equation*}
\ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string}
  \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{equation*}
%
The expression queries a database and gets back a list of news values as the result. Recall from the
earlier discussion that there are two contextual information that are desirable to track for this
expression. First, the call to the \ident{query} primitive requires \emph{database access}. Second,
the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
an unsecure network connection.
The \emph{coeffect systems} developed in this thesis capture this information in the following way:
%
\begin{equation*}
\begin{array}{l}
 \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
   \ident{cmd}\!:\!\ident{string})}{ \cclrd{\{ \ident{database} \}}}  \\[-0.1em]
 \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{array}
\end{equation*}
%
Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
context $\Gamma$. In other words, coeffect systems do not keep track just of the variables available
in the context -- they also capture detailed information about the execution environment.
In the above example, the system tracks meta-data about the variables and annotates \ident{password} 
as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
that the execution requires an access to database.

@@

It is well-known fact that \emph{effects} correspond to \emph{monads} and languages such as Haskell 
use monads to provide a limited form of effect system. An interesting observation made in this thesis 
is that \emph{coeffects}, or systems for tracking contextual information, correspond to the category 
theoretical dual of monads called \emph{comonads}. The details are explained when discussing the 
semantics of coeffects throughout the thesis.

@@

\begin{itemize}
\item In Chapter~\ref{ch:pathways} we give an overview of related work. Most importantly, we show
  that the idea of context-aware computations can be naturally approached from a number of
  directions developed recently in theories of programming languages (including type and effect
  systems, categorical semantics and sub-structural logics).

@@

\item Related work is presented in Chapter~\ref{ch:pathways} and throughout the thesis. 
  One important direction deserves further exploration, and so Chapter~\ref{ch:conclusions}
  starts with a brief discussion of a different approach to tracking contextual information that 
  arises from modal logics. Finally, the rest of Chapter~\ref{ch:conclusions} discusses approaches 
  for implementing the presented theory in main-stream programming languages and concludes.
\end{itemize}

>>>>

\item \textbf{Resources \& data availability.} When creating a mobile application, the program may
  (or may not) be granted access to device capabilities such as GPS sensor, social updates or battery
  status. We would like to know which of the capabilities are required and which are optional
  (\ie~enhance the user experience, but there is a fallback strategy). Equally, on the server-side,
  we might have access to different database tables, depending on the role of the user.
\end{itemize}

\noindent
Most developers do not perceive the above as programming language flaws. They are simply common
programming problems (at most somewhat annoying and tedious) that have to be solved. However, this
is because it is not apparent that a suitable language extension could make the above problems
significantly easier to solve. As the number of distinct contexts and their diversity increases,
these problems will become even more commonplace.

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
\kvd{for}~\ident{header},\ident{value}~\kvd{in}~\ident{header}~\kvd{do}\\[-0.25em]
\hspace{2em}\kvd{match}~\ident{header}~\kvd{with}\\[-0.25em]
\hspace{2em}|~\str{accept} \rightarrow~ \ident{req.Accept} \leftarrow \ident{value}\\[-0.25em]
%
\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_USERAGENT}\\[-0.25em]
\hspace{2em}|~\str{user-agent} \rightarrow~ \ident{req.UserAgent} \leftarrow \ident{value}\\[-0.25em]
\prepk{\#else}\\[-0.25em]
\hspace{2em}|~\str{user-agent} \rightarrow~
  \ident{req.Headers}.[\,\ident{HttpHeader.UserAgent}\,] \leftarrow \ident{value}\\[-0.25em]
\prepk{\#endif}\\[-0.25em]
%
\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_REFERER}\\[-0.25em]
\hspace{2em}|~\str{referer} \rightarrow~ \ident{req.Referer} \leftarrow \ident{value}\\[-0.25em]
\prepk{\#else}\\[-0.25em]
\hspace{2em}|~\str{referer} \rightarrow~
  \ident{req.Headers}.[\,\ident{HttpHeader.Referer}\,] \leftarrow \ident{value}\\[-0.25em]
\prepk{\#endif}\\[-0.25em]
%
\hspace{2em}|~\ident{other} \rightarrow~ \ident{req.Headers}.[\,\ident{other}\,] \leftarrow \ident{value}\\[-0.25em]
\end{array}
\end{equation*}

\caption{Conditional compilation in the HTTP module of the F\# Data library}
\label{fig:introduction-context-http}
\end{figure}

@@

As an alternative to conditional compilation, developers can use dynamic loading. For example,
on Android, programs can access API from higher level platform dynamically using techniques like
reflection and writing wrappers. This is even more error prone. As noted in an article\footnote{Retrieved 
from: \url{http://android-developers.blogspot.com/2009/04/backward-}\\\url{compatibility-for-android.html}}
introducing the technique ``Remember the mantra: if you haven't tried it, it doesn't work.'' 
Again, it would be reasonable to expect that statically-typed languages can provide a better solution.

@@

Another example related to the previous one is when libraries use meta-programming techniques,
such as LINQ \cite{app-linq,app-linq-essence} or F\# quotations \cite{app-fsharp-metaprog}, to 
translate code written in a subset of a host language to some other target language, such as SQL, 
OpenCL or JavaScript. For database access, this is a recently developed technique replacing embedded 
SQL discussed in the introduction, but it is a more broadly applicable technique for programming
in heterogeneous environments. It lets developers targets multiple runtimes that have limited 
execution capabilities.

@@

When cross-compiling to a limited environment, we cannot always support the full source language.
The example with LINQ and SQL demonstrates the importance of this problem. As of March 2014,
Google search returns 11,800 results for the message above and even more results (44,100) for a
LINQ error message \emph{``Method X has no supported translation to SQL''} caused by a similar
limitation.

@@

The previous two examples were related to the non-existence of some library functions in a different
execution environment. Another common factor was that they were related to the execution context of 
the whole program or a function scope. However, contextual properties can also be associated with
specific variables.

For example, consider the following code sample that accesses a database by building a SQL query
using string concatenation. For the purpose of the demonstration, this example does not use LINQ, 
but an older approach with a parameterized SQL query written as a string:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{query} = \ident{sprintf}~\str{SELECT * FROM Products WHERE Name='\%s'}~\ident{name}\\[-0.25em]
\kvd{let}~\ident{cmd} = \kvd{new}~\ident{SqlCommand}(\ident{query})\\[-0.25em]
\kvd{let}~\ident{reader} = \ident{cmd.ExecuteReader}()
\end{array}
\end{equation*}
%
The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
\cite{app-sql-injection}. An attacker could enter \str{'; DROP TABLE Products -{}-} as their 
\ident{name} and delete the database table ``Products''! For this reason, most libraries discourage 
building SQL commands by string concatenation, but there are still many systems that do so.

Again, this example demonstrates a more general property. Sometimes, it is desirable to track additional
meta-data about variables that are in some ways special. Such meta-data can determine how the variables
can be used. Here, \ident{name} comes from the user input. This information about the value should
be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
not directly contain user input (in an unchecked form).

@@

In the security context, such marking of values (but at run-time) is called \emph{tainting} 
\cite{app-tainting-sql}, but the technique is a special case of more general \emph{provenance}
tracking \cite{app-provenance-future}. This can be useful when working with data in other contexts. 
For example, data jounralsts might want to propagate meta-data about the quality and the 
information source -- is the source trustworthy? Is the data up-to-date? Such meta-data could 
propagate to the result and tell us important information about the calculated results.

@@

The information about access patterns can be used to efficiently compile the computation
by preallocating the necessary space (as we know which sub-range of the array might be accessed).
It also allows better handling of boundaries \cite{app-stencil-dsl}. For example, to simplify 
wrap-around behaviour we could pad the \ident{input} with a known number of elements from the 
other side of the array.

@@

The four examples presented in the previous section cover different kinds of \emph{context}.
The context includes notions such as execution environment, capabilities provided by the 
environment or input and meta-data about the input and variables through which it is accessed.

@@

\caption{Client/server news reader app implemented in a context-aware language}
\label{fig:introduction-sample}
\end{figure}

\newcommand{\introfuncblack}[3]{ \ident{#1 \hspace{-1em}\footnotesize{@} \{ #2 \}} $\rightarrow$ \ident{#3}  }
\newcommand{\introfunc}[3]{ \ident{#1 \hspace{-1em}\footnotesize{@} \aclrd{\{ #2 \}}} $\rightarrow$ \ident{#3}  }
\newcommand{\introvar}[2]{ \ident{#1 \hspace{-1em}\footnotesize{@} \cclrd{#2} } }

@@

The approach advocated in this thesis is to track information about context requirements using
the type system. To make this practical, the system should also provide at least a partial support for
automatic type inference, as the information about context requirements makes the types more complex.
An inspiring example might be the F\# support for units of measure \cite{app-units} -- the user has
to explicitly annotate constants, but the rest of the information is inferred automatically.

Furthermore, integrating contextual information into the type system can provide information for
modern developer tools. For example, many editors for F\# display inferred types when placing mouse
pointer over an identifier. For \ident{fetchLocalNews}, the tip could appear as follows:
%
\definecolor{tooltipy}{rgb}{0.98,0.97,0.76}
\definecolor{tooltipb}{rgb}{0.7,0.7,0.7}
\vspace{0.5em}
\begin{equation*}
\hspace{5em}
\begin{array}{l}
\ident{\textbf{fetchLocalNews}}\\
\hspace{1em} \fcolorbox{tooltipb}{tooltipy}{
	\introfuncblack{unit}{gps, rpc}{(news~list)~async}
}
\end{array}
\end{equation*}
\vspace{0.5em}

@@

\item The \ident{fetchNews} function requires database access and so it can only run on the server-side
  (or on a thick client with a local copy of the database, such as a desktop computer with an offline mode).

@@

In this scenario, the main benefit of using an integrated context-aware language would be the
ability to design appropriate abstractions using standard mechanisms of the language. For
cross-compilation, we can structure code using functions, rather than relying on \prepk{\#if}
directives. Similarly, the splitting between client-side, server-side and shared code can be
done using ordinary functions and modules (with shared functions reused) -- rather than having 
to split the application into separate independent libraries or projects.

@@

\paragraph{Type systems.}
A type system is a form of static analysis that is usually specified by \emph{typing judgements}
such as $\Gamma \vdash e : \tau$. The judgement specifies that, given some variables described by
the context $\Gamma$, the expression $e$ has a type $\tau$.
The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an
expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int},
\ident{float}, or even \ident{string}, depending on the types of the variables. For example, 
the following is a valid typing judgement in F\# \cite{app-fsharp}:
%
\begin{equation*}
x\!:\!\ident{int},~y\!:\!\ident{int} \vdash x+y : \ident{int}
\end{equation*}
%
This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
be \ident{int}. In F\#, the expression would also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
but not, for example, in a context where $x$ has a type \ident{int} and $y$ has a type \ident{string}.

@@

The systems for tracking of context requirements developed in this thesis are inspired by the idea
of effect systems. To demonstrate our approach, consider the following call from the sample program
shown earlier -- first using standard ML-like type system:
%
\begin{equation*}
\ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string}
  \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{equation*}
%
The expression queries a database and gets back a list of news values as the result. Recall from the
earlier discussion that there are two contextual information that are desirable to track for this
expression. First, the call to the \ident{query} primitive requires \emph{database access}. Second,
the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
an unsecure network connection. The \emph{coeffect systems} developed in this thesis capture this 
information in the following way (we slightly refine the notation later):
%
\begin{equation*}
\begin{array}{l}
 \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database} \}}}  \\[-0.1em]
 \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{array}
\end{equation*}
%
Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
context $\Gamma$. In other words, coeffect systems do not keep track just of the variables available
in the context -- they also capture detailed information about the execution environment.
In the above example, the system tracks meta-data about the variables and annotates \ident{password} 
as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
that the execution requires an access to database.

@@

It is well-known fact that \emph{effects} correspond to \emph{monads} and languages such as Haskell 
use monads to provide a limited form of effect system. An interesting observation made in this thesis 
is that \emph{coeffects}, or systems for tracking contextual information, correspond to the category 
theoretical dual of monads called \emph{comonads}. The details are explained throughout the thesis.

@@

\begin{itemize}
\item In Chapter~\ref{ch:pathways} we give an overview of related work. Most importantly, we show
  that the idea of context-aware computations can be naturally approached from a number of
  directions developed recently in the theory of programming languages (including type and effect
  systems, categorical semantics and sub-structural logics).

@@

\item Related work is presented in Chapter~\ref{ch:pathways} and throughout the thesis. 
  One important direction deserves further exploration, and so Chapter~\ref{ch:unified}
  also includes a brief discussion of a different approach to tracking contextual information that 
  arises from modal logics. This is followed by an outline of the future work that is needed
  in order to implement the presented theory in main-stream programming languages.
\end{itemize}
